The Unicode Cookbook \linebreak for Linguists

### 2018-03-06 book/chapters/orthography_profiles.tex
these changes almost complete the remaining reviews

Given these goals, Unicode locale descriptions (see
Section~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
such locale descriptions in the daily linguistic practice, namely: (i) the
XML-structure is too verbose to easily and quickly produce or correct manually,
(ii) locale descriptions are designed for a wide scope on information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct locale description for a specific source at hand,
then it is nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language.\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{rule-ordering}
).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Section~\ref{use-cases} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} 
	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
       includes the information pertinent to the orthography. 
	\item \textsc{A profile is a tab-separated CSV file with an obligatory header
       line}. A minimal profile can have just a single column, in which case
       there will of course be no tabs, but the first line will still be the
       header. For all columns we assume the name in the header of the CSV file
       to be crucial. The actual ordering of the columns is unimportant. Because 
       tabs and newlines are field separators of the profile, these characters 
       will lead to problems when they occur in the original data.\footnote{The main
       reason to choose for tab-delimited, and not for quoted comma-separated files
       is that we want the profiles to be easily manipulated by hand. Tab-separation 
       is strongly preferred for this reason.}
	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
       \textsc{tag: value} format. Metadata about the orthographic description
       given in the orthography profile includes, minimally, (i) author, (ii)
       date, (iii) title of the profile, (iv) a stable language identifier
       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Further, (vi) the tokenization method and (vii)
       the unicode normalisation used should be documented here (see below).
    	\item \textsc{Separate lines with comments are not allowed}. Comments that
       belong to specific lines will have to be put in a separate column of
       the CSV file, e.g.~add a column called \textsc{comments}.            
\end{enumerate}

>>>>

Given these goals, Unicode locale descriptions (see
Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
such locale descriptions in the daily linguistic practice, namely: (i) the
XML-structure is too verbose to easily and quickly produce or correct manually,
(ii) locale descriptions are designed for a wide scope on information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct locale description for a specific source at hand,
then it is nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}
).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Section~\ref{r-implementation} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} 
	\item \textsc{A profile is a unicode UTF-8 encoded text file} 
	   normalized following NFC (or NFD if specified in the metadata)
       that includes the information pertinent to the orthography.\footnote{See 
	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
	   no-BOM and LF line breaks because of the pitfalls they avoid. 
	   A keen reviewer notes, however, that specifying 
	   a convention for line endings and BOM is overly strict because most 
	   computing environments (now) transparently handle both alternatives. 
	   For example, using Python a file can be decoded using the encoding 
	   ``utf-8-sig'', which strips away the BOM (if present) and reads 
	   an input full in text mode, so that both line feed variants ``LF'' and 
	   ``CRLF'' will be stripped.}

%	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
%       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
%       includes the information pertinent to the orthography. 
	\item \textsc{A profile is a CSV file with an obligatory header
       line}. A minimal profile must have a single column with the header ``Grapheme''. 
	   For any additional columns, the name in the header of the CSV file
       must be specified. The actual ordering of the columns is unimportant. Cells should 
	   not contain tabs or newlines. The header list must be delimited 
	   in the same way as the rest of the file's contents. Each record must be kept 
	   on a separate line. Separate lines with comments are not allowed. Comments that
       belong to specific lines must be put in a separate column of
       the CSV file, e.g.~add a column called \textsc{comments}.

\begin{comment}	   
	\item \textsc{Metadata should be added in a separate UTF-8 text file} using a basic
       \textsc{tag: value} format. Metadata about the orthographic description
       given in the orthography profile includes, minimally, (i) author, (ii)
       date, (iii) title of the profile, (iv) a stable language identifier
       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Further, (vi) the tokenization method and (vii)
       the unicode normalization used should be documented here (see below).
\end{comment}

	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
	   the JSON-DL dialect specified in \textit{Metadata Vocabulary for Tabular 
	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
	   format allows easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
	   which should be used to specify information about the orthographic description 
	   in the orthography profile. The orthography profile metadata should minimally include: 
	   (i) author, (ii) date, (iii) title of the profile, (iv) a stable language identifier
       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Further, (vi) the tokenization method and (vii)
       the Unicode normalization form used should be documented here (see below).
	   In accordance with the \textit{Metadata Vocabulary for Tabular 
	   Data}, the metadata's filename must consist of the orthography 
	   profile's filename appended with ``-metadata.json''.\footnote{JSON-DL metadata 
	   is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, 
	   see: \url{http://cldf.clld.org/}.}

\end{enumerate}

### 2018-03-06 book/chapters/pitfalls.tex
these changes almost complete the remaining reviews

The Unicode Standard deals with complex graphemes only inasmuch they consist of
base characters with diacritics (see
Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
clusters}. Complex graphemes consisting of multiple base characters,
like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
Section~\ref{the-unicode-approach}).

@@

The actual intended meaning of a Unicode codepoint is a combination of the
name, the block and the character properties (see
Section~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
codepoint of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are avaialbe at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

@@

@@

>>>>

The Unicode Standard deals with complex graphemes only inasmuch they consist of
base characters with diacritics (see
Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
clusters}. Complex graphemes consisting of multiple base characters,
like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
Chapter~\ref{the-unicode-approach}).

@@

The actual intended meaning of a Unicode codepoint is a combination of the
name, the block and the character properties (see
Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
codepoint of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are available at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

@@

% ==========================
\section{Pitfall: Encodings}
\label{encoding}
% ==========================

% Section "Pitfall: File formats" may profit most from a more transparent terminology, distinguishing levels of encoding, rather than talking about how texts "appear inside some kind of computer file".

Before we discuss the pitfall of different file formats in Section \ref{pitfall-file-formats}, it is pertinent to point out that the common usage of the term \textsc{encoding}, unfortunately, does not distinguish between \textit{encoded} sequences of code points and text \textit{encoded} as bit patterns. Recall, a code point is simply a numerical representation of some defined entity; or in other words, a code point is a character encoding-specific unique identifier or ID. In the Unicode Standard encoding, code points are numbers that serve as unique identifiers, each of which is associated with a set of character properties defined by the Unicode Consortium in the Unicode Character Database.\footnote{\url{https://www.unicode.org/ucd/}} The number of each code point can be \textit{encoded} in various formats, including as a decimal integer (e.g.\ 112), as an 8-bit binary sequence (01110000) or hexadecimal (0070). This example Unicode code point, \uni{0070}, represents \textsc{latin small letter p} and its associated Unicode properties, such as it belongs to the category Letter, Lowercase [Ll], in the Basic Latin block, and that its title case and upper case is associated with code point \uni{0050}.\footnote{See also Chapter \ref{the-unicode-approach}.}

% 0070;LATIN SMALL LETTER P;Ll;0;L;;;;;N;;;0050;;0050

The other meaning of encoding has to do with the fact that computers represent data and instructions in patterns of bits. A bit pattern is a combination of binary digits arranged in a sequence. And how these sequences are carved up into bit patterns is determined by how they are \textit{encoded}. Thus the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems because they determine how sequences of bit patterns are determined, which are then mapped to characters. The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e. the code point, which can be encoded differently by different encoding systems.

% Characters encoded, but not seen.
Note also that all computer character encodings include so-called \textsc{control characters}, which are non-printable sometimes action-inducing characters, such as the null character, bell code, backspace, escape, delete, and line feed. Control-characters can interact with encoding schemes. For example, some programming languages make use of the null character to mark the end of a string. Line breaks are part of the text, and as such as covered by the Unicode Standard. But they can be problematic because line breaks differs from operating system-to-operating system in how they are encoded. These variants are discussed in Section \ref{pitfall-file-formats}.

% This section should also mention that line breaks are actually part of the text, and as such also covered by the Unicode standard. Again regarding line breaks, the recommendation "that everybody use this encoding [only LF] whenever possible" (page 31) seems less well motivated (and specific) than most other recommendations in the book. Some common formats, e.g. CSV (as specified by  RFC 4180 [4]), specify "CRLF" as line break. Thus, the perceived "strong tendency" towards "LF" may be just that.

@@

% ==========================
\section{Pitfall: Incomplete implementations}
\label{incomplete-implementations}
% ==========================
Another pitfall that we encounter when using the Unicode Standard is its incomplete implementation in different standards and programming languages, e.g.\ SQL, XML, XLST, Python. For example, although the Unicode Standard mandates that the comparison of Unicode text be done using normalized text, this is not the case with the equality operator ``=='' in Python. Furthermore, it is not always transparent what the operating system or specific software applications do when text is being copied and pasted. For example, copy and pasting the character sequence \uni{0061} \textsc{latin small letter a} <a> and \uni{0301} \textsc{combining acute accent} <\dia{0301}>, visually <á>, into the text editor TextWrangler leaves the sequence decomposed as two characters. But when pasting the decomposed sequence into RStudio, and other software programs, the sequence becomes precomposed as \uni{00E1} \textsc{latin small letter a with acute}, i.e.\ <á>.

% One Unicode pitfall which may be worth adding is the problem with incomplete implementations of the Unicode Standard. This is a problem most big-enough standards suffer from, e.g. SQL, XML, XSLT. So although the Unicode Standard mandates that comparison of Unicode text should always be done using normalized text, this is not how the equality operator "==" is implemented for Unicode text in the Python programming language.

### 2018-03-06 book/chapters/unicode_approach.tex
these changes almost complete the remaining reviews

\section{Character encoding system}

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting. Underlyingly, the character encoding is
represented by a range of numerical values called a \textsc{code space}, which
is used to encode a set of characters. A \textsc{code point} is a unique
non-negative integer within a code space (i.e.~within a certain numerical
range). In the Unicode Standard character encoding system, an \textsc{abstract
character}, for example the \textsc{latin small letter p}, is mapped to a
particular code point, in this case the decimal value 112, normally represented in
hexadecimal, which then looks in Unicode parlance as
\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
computers will of course treat this code point binary (base-2) as 11100000, as
can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
16 = 112$.} That encoded abstract character is rendered on a computer screen (or
printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
the context in which that character appears.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. So, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed code point}, \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code point
character sequences. The problem that there multiple encodings possible for the
same text element has been acknowledged early on in the Unicode Standard
(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
treated identically to the precomposed \uni{00F1}), and a system of
\textsc{canonical equivalence} is available for such situations. Basically, the
Unicode Standard offers different kind of normalizations to either decompose all
precomposed characters (called \textsc{NFD}, \textsc{normalization form
canonical decomposition}), or compose as much as possible combinations (called
\textsc{NFC}, \textsc{normalization form canonical composition}). In current
practice of software development, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form.

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. The Unicode
Standard provides technological specifications for creating locale specific data
in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
that defines a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, there are various drawbacks of locale
descriptions for the daily practice of linguistic work in a multilingual setting.

>>>>

\section{Character encoding system}
\label{character-encoding-system}

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems. The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e. the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
represented by a range of numerical values called a \textsc{code space}, which
is used to encode a set of characters. A \textsc{code point} is a unique
non-negative integer within a code space (i.e.~within a certain numerical
range). In the Unicode Standard character encoding system, an \textsc{abstract
character}, for example the \textsc{latin small letter p}, is mapped to a
particular code point, in this case the decimal value 112, normally represented in
hexadecimal, which then looks in Unicode parlance as
\uni{0070}.
%\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
%(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
%(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
%computers may treat this code point as an 8-bit binary (base-2) sequence (11100000), as
%can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
%(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
%16 = 112$.} 
That encoded abstract character is rendered on a computer screen (or
printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
the context in which that character appears.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. So, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed code point}, \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code point
character sequences. The problem that there multiple encodings possible for the
same text element has been acknowledged early on in the Unicode Standard
(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
treated identically to the precomposed \uni{00F1}), and a system of
\textsc{canonical equivalence} is available for such situations. Basically, the
Unicode Standard offers different kind of normalizations to either decompose all
precomposed characters (called \textsc{NFD}, \textsc{normalization form
canonical decomposition}), or compose as much as possible combinations (called
\textsc{NFC}, \textsc{normalization form canonical composition}).\footnote{See the 
Unicode Standard Annex \#15, Unicode Normalization Forms (\url{http://unicode.org/reports/tr15/}), 
which provides a detailed description of normalization algorithms and illustrated examples.} 
In current practice of software development, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form. We discuss Unicode Normalization 
in detail in Section \ref{pitfall-canonical-equivalence}.

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. They are also 
grapheme clusters in these orthographies for processes such as collation.\footnote{\url{https://www.unicode.org/glossary/#collation}}

The Unicode Standard provides technological specifications for creating locale specific data
in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
that defines a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, there are various drawbacks of locale
descriptions for the daily practice of linguistic work in a multilingual setting.
